Bayesian methods for black hole merger parameter estimation with LISA.

### 2019-04-04 paper/paper_PE.tex
Work on the paper

\subsection{The LISA response - OLD}
\label{sec:lisaresponse}

\SM{------------------------------------------------------}

\SM{[copy-paste from previous paper -- just here to copy needed formulas / refs]}

We begin by detailing the model that we use for the response of a LISA-like detector, together with the assumptions used and their limitations. Since our aim is to assess the accuracy of our direct Fourier-domain treatment of the response, we can focus on the gravitational-wave contribution to the basic single-link observables. We therefore use a somewhat simplified model for the time-domain response~\cite{Krolak+04}, ignoring corrections that would be crucial from the point of view of noise cancellations, keeping in mind that the response model can be enriched later without affecting the conclusions of the present analysis.

The frequency-shift response for a single link~\eqref{eq:yslr} was derived in Ref.~\cite{EW75} (see also~\cite{CR02, RCP04, Finn08, Cornish09}). Several assumptions enter the result as written in~\eqref{eq:yslr}: (i) effects of the order $v/c$ are neglected, including for instance the special relativistic Doppler effect created by the relative speeds of the spacecraft on their orbits (ii) the propagation is assumed to take place in a flat spacetime, perturbed only by the gravitational wave; thus the gravitational redshift as well as the deflection of light created by the gravitational potential of the Sun is ignored (iii) all geometric factors are evaluated at a single time, whereas one should consider the beam as propagating from the position of the first spacecraft at the time of emission to the position of the second scapecraft at the time of reception, leading to a point-ahead effect.

Additionally, we limit ourselves to a rigid model for the orbits of the constellation, namely we assume that the constellation remains in an equilateral configuration with fixed armlengths. These simplified orbits neglect (iv) effects of order $e^{2}$ from the eccentricity of the individual Keplerian orbits, (v) the effect of gravitational perturbations coming from other celestial bodies, such as the Earth, the quadrupole of the Sun, and the other planets. Note that although we can neglect all the effects (i)-(v) for our present study, keeping track of these corrections is crucial for the purpose of laser noise cancellations, and led to the development of new generations of TDI observables~\cite{Tintoliving}.

It is natural to split the response~\eqref{eq:yslr} into two steps: first the orbital delay related to the orbit around the Sun of the whole constellation, and then the constellation response. The baselines for the delays are indeed very different in the two cases. Geometrical projection factors aside, we have for the orbit around the Sun $R=1\text{au}=1.5\times 10^{8} \text{km}$, while the detector armlength is $L=2.5\times 10^{6}\text{km}$ (in the configuration proposed in~\cite{LISA17}). It is natural to define two transfer frequencies for the two relevant length scales for the delays, defined such that a wavelength fits within this length scale, i.e. $2\pi f d = 1$. This gives
\begin{subequations}\label{eq:transferfrequencies}
\begin{align}
	f_{R} &= 3.2\times10^{-4}\Hz \,,\\
	f_{L} &= 1.9\times 10^{-2}\Hz \,.
\end{align}
\end{subequations}
The LISA response will behave qualitatively differently on the three frequency bands $f \leq f_{R}$, $f_{R} \leq f \leq f_{L}$ and $f_{L} \leq f$.

The first stage of the response, the orbital delay, consists simply in applying the varying time delay to bring the wavefront sampling point from the SSB reference to the center of the LISA triangular constellation, common to all $y_{slr}$ observables. For $h^{\rm TT}$ the transverse-traceless gravitational waveform in matrix form, we write this orbital time delay as
\be\label{eq:defresponse0}
	h_{0}^{\rm TT} (t) = h^{\rm TT}(t-\hatk\cdot p_{0}) \,,
\ee
with $p_{0}$ the position of the constellation center, which follows the Earth orbit around the Sun. The second stage of the response calculation comprises the remaining, constellation-centered response. For the single-link contribution to the response, for the laser link from spacecraft $s$ to spacecraft $r$ along a path in direction $n_l$, we write
\begin{align}\label{eq:defresponseL}
	y_{slr} &= \frac{1}{2} \frac{1}{1 - \hatk\cdot n_{l}} \nn\\
	& \cdot n_{l}\cdot \left[ h_{0}^{\rm TT}(t - L - \hatk\cdot p^{L}_{s}) - h_{0}^{\rm TT}(t - \hatk\cdot p^{L}_{r}) \right] \cdot n_{l}\,,
\end{align}
where we reference the positions of the spacecraft relative to the center of the constellation, $p^{L}_{A} \equiv p_{A} - p_{0}$.

As described in App.~\ref{app:notation} We will decompose the full signal in the contributions of the individual spin-weighted spherical modes $h_{\ell m}$, whose Fourier transforms are assumed to have a smooth amplitude and phase. First, we define the matrices $P_{+},P_{\times}$ such that, in the sense of matrices,
\be
	h^{\rm TT} = h_{+}P_{+} + h_{\times}P_{\times} \,.
\ee
We focus only on positive frequencies. Assuming that the approximation~\eqref{eq:zeronegativef} applies, we consider a single mode contribution, $h=h_{\ell m}$ with $m>0$. For each given mode we define a complex matrix $P_{\ell m}$ incorporating the spin-weighted spherical harmonic constant factor as
\be
	P_{\ell m} =
	\begin{cases}
	\frac{1}{2} {}_{-2}Y_{\ell m} \left( P_{+} + i P_{\times} \right) \text{ for } m>0\,,\\
	\frac{1}{2} {}_{-2}Y_{\ell m}^{*} \left( P_{+} - i P_{\times} \right) \text{ for } m<0\,.
	\end{cases}
\ee

We now turn to the transformation of~\eqref{eq:defresponse0} and~\eqref{eq:defresponseL} to the Fourier domain. Applying a pure delay as in~\eqref{eq:defresponse0} translates into
\be\label{eq:G0}
	G_{0}(f, t) = e^{-2i\pi f d_{0}(t)} \,,
\ee
with $d_{0} = -\hatk \cdot p_{0}$ the delay associated to the orbit around the Sun. For the leading-order response~\eqref{eq:transferlocal}, this gives a Fourier-domain transfer function common to all modes, that is a pure phase factor, proportional to the frequency but also $t_{f}$-dependent:
\be\label{eq:transfer0local}
	\calT_{0}^{\rm local}(f) = G_{0}(f, \tf)\,.
\ee
If $(\lambda, \beta)$ are the ecliptic longitude and latitude of the source in the sky, and if the orbital phase is set by convention to $0$ at $t=0$, the orbital delay has the simple expression
\be\label{eq:delay0}
	d_{0}(t) = -R \cos\beta \cos\left(\Omega_{0}t - \lambda\right)\,.
\ee

For the constellation part of response~\eqref{eq:defresponseL}, treated separately from the delay~\eqref{eq:delay0}, we write
\begin{align}\label{eq:decomposeGslr}
	F_{slr}^{L}(t) &= \frac{1}{2} \frac{1}{1 - \hatk\cdot n_{l}(t)} n_{l}(t) \cdot P_{\ell m} \cdot n_{l} (t) \,,\nn\\
	d_{s}(t) &= - k\cdot p_{s}^{L}(t) \,, \quad d_{r}(t) = - k\cdot p_{r}^{L}(t) \,,\nn\\
	G_{slr}^{L}(f,t) &=  F_{slr}^{L}(t) \left( e^{-2i\pi f (d_{s,L}(t) + L)} - e^{-2i\pi f d_{r}(t)} \right) \,.
\end{align}
The superscript $L$ indicates that the orbital delay~\eqref{eq:delay0} is not included. Since we also assume the rigid approximation for the constellation, where the armlengths are fixed, a particular simplification occurs when combining these individual delays, thanks to the relation $p^{L}_{r} - p^{L}_{s} =  L n_{l}$:
\begin{align}\label{eq:GslrL}
	G_{slr}^{L}(f,t) &= \frac{i \pi f L}{2} \sinc \left[ \pi f L\left(1-\hatk\cdot n_{l} \right) \right] \nn\\
	& \quad \cdot \exp\left[ i \pi f \left( L + \hatk\cdot \left( p_{1}^{L} + p_{2}^{L} \right) \right) \right]  n_{l} \cdot P_{\ell m} \cdot n_{l} \,,
\end{align}
with all time-dependent vectors evaluated at $t$. This expression is well known as describing the frequency-dependency in the LISA response~\cite{Larson+99, Cornish01, CR02, RCP04}. In the local approximation~\eqref{eq:transferlocal}, the Fourier-domain transfer function then reads
\begin{align}\label{eq:transferLlocal}
	\calT_{slr}^{L, \mathrm{local}}(f) &= G_{slr}^{L}(f, \tf) \,.
\end{align}
For plotting purposes, we will also define
\be\label{eq:transferLenvelope}
	\overline{\calT}_{slr}^{L} (f) = \frac{i \pi f L}{2} n_{l} \cdot P_{\ell m} \cdot n_{l} (\tf)
\ee
which will serve as an estimate for the enveloppe function of the response, devoid of the zero-crossings at high frequencies of the $\sinc$ term in~\eqref{eq:GslrL}.

Note that, if the corrections of Sec.~\ref{subsec:delays}, for non-negligible $\dot d$, are included for the constellation delays, the transfer function will not have this simple form anymore, as $\dot{d}$ will have a different velocity-dependent expression for the sending and receiving spacecraft. One must then separately handle $d_{s}$ and $d_{r}$ in~\eqref{eq:decomposeGslr} to compute the corrections.

The orbital response~\eqref{eq:transfer0local}-\eqref{eq:delay0} takes a simple analytic form, but the phase contribution of this delay is significant across most of the frequency band and can be large for $f \gg f_{R}$.

The constellation response~\eqref{eq:GslrL}-\eqref{eq:transferLlocal} can be interpreted as the Fourier-domain translation of a discrete derivative taken on the waveform. The leading factor in~\eqref{eq:GslrL} shows that the amplitude of the response is proportional to $f$ in the low-frequency limit $f\ll f_{L}$, where the other factors are essentially unity. For $f\gtrsim f_{L}$, the $\sinc$ and the phase of the exponential generate additional structure in the response, including zero-crossings when the projected armlength is an integer number of wavelengths. From~\eqref{eq:transferLlocal}, an expansion for small $f\ll f_{L}$ yields back a Fourier-domain analog of the low-frequency approximation of the response~\cite{Cutler97, RCP04}, which is equivalent to having two LIGO-type interferometers turned by $\pi/4$ and set in motion.

For analysis of the response we need a concrete set of gravitational waveforms. We will use the PhenomD model~\cite{Khan+15,Husa+15}, which provides Fourier-domain inspiral-merger-ringdown waveforms for aligned spins. We refer to App.~\ref{app:precLISA} for a brief discussion of the prospects for applying our formalism for the LISA response to precessing Fourier-domain waveforms.

\SM{------------------------------------------------------}

In our rigid approximation, delays are simply all constant and constant to $L$. Using the shortcut notation $y_{sr,L} = y_{slr}(t - L)$, the first-generation TDI Michelson observable $X$ reads
\begin{align}
	X &= y_{31} + y_{13,L} + \left( y_{21} + y_{12,L} \right)_{2L} \nn\\
	& - \left( y_{21} + y_{12,L} \right) - \left( y_{31} + y_{13,L} \right)_{2L} \,,
\end{align}
with the other Michelson observables $Y$, $Z$ being obtained by cyclic permutation. The independent combinations $A$, $E$ and $T$ are then defined as
\bsub
\begin{align}
	A &= \frac{1}{\sqrt{2}} \left( Z - X \right) \,,\\
	E &= \frac{1}{\sqrt{6}} \left( X - 2Y + Z \right) \,,\\
	T &= \frac{1}{\sqrt{3}} \left( X + Y + Z \right) \,.
\end{align}
\esub
These channels are independent under the approximation of an uncorrrelated noise between the arms. With constant delays in the rigid approximation, the TDI combinations take the form
\bsub\label{eq:aet}
\begin{align}
	\tilde{a} &= \left[ (1+z) \left( \tilde{y}_{31} + \tilde{y}_{13} \right) - \tilde{y}_{23} - z \tilde{y}_{32} - \tilde{y}_{21} - z \tilde{y}_{12} \right] \,,\\
	\tilde{e} &= \frac{1}{\sqrt{3}} \left[ (1-z)\left( \tilde{y}_{13} - \tilde{y}_{31} \right) + (2+z) \left( \tilde{y}_{12} - \tilde{y}_{32} \right) \right. \nn\\
	&\qquad\quad \left. + (1+2z) \left( \tilde{y}_{21} - \tilde{y}_{23} \right) \right] \,,\\
	\tilde{t} &= \frac{\sqrt{2}}{\sqrt{3}} \left[ \tilde{y}_{21} - \tilde{y}_{12} + \tilde{y}_{32} - \tilde{y}_{23} + \tilde{y}_{13} - \tilde{y}_{31} \right] \,,
\end{align}
\esub
where we have introduced the rescalings
\bsub\label{eq:scalingAET}
\begin{align}
	\tilde{a}, \tilde{e} &= \frac{1}{i \sqrt{2} z \sin 2\pi f L}\times \tilde{A}, \tilde{E} \,,\\
	\tilde{t} &= \frac{1}{2\sqrt{2} z^{3} \sin \pi f L \sin 2\pi f L} \times \tilde{T} \,.
\end{align}
\esub
Scaling out the same square factors from the noise power spectral density as
\bsub
\begin{align}
	S_{n}^{A}, S_{n}^{E} &= 2 \sin^{2} 2\pi f L \times S_{n}^{a}, S_{n}^{e} \,,\\
	S_{n}^{T} &= 8 \sin^{2} \pi f L \sin^{2} 2\pi f L S_{n}^{t} \,,
\end{align}
\esub
the reduced PSD for the three channels take the form
\bsub
\begin{align}
	S_{n}^{a} = S_{n}^{e} &= 2 \left( 3 + 2\cos 2\pi f L + \cos 4 \pi f L \right) S^{\rm pm}(f) \nn\\
	& \quad + \left( 2 + \cos 2\pi f L \right) S^{\rm op}(f) \,,\\
	S_{n}^{t} &= 4 \sin^{2} 2\pi f L S^{\rm pm}(f) + S^{\rm op}(f) \,,
\end{align}
\esub
with $S^{\rm pm}$ the test-mass noise PSD and $S^{\rm op}$ the optical noise PSD. Here we ignore the confusion noise coming from the background of galactic binaries in the  LISA band. Note that the prefactors~\eqref{eq:scalingAET} are oscillatory and have zero-crossings at high frequencies, and scaling them out from both the signal and the noise is an approximation, as we can expect that the real delays would lead to imperfect cancellations in the vicinity of the zero-crossings.

@@

Introducing the standard matched-filter inner product as
\be
	\left( \tilde{a} | \tilde{b} \right) = 4 \mathrm{Re} \int_{0}^{+\infty} df \; \frac{\tilde{a}(f) \tilde{b}^{*}(f)}{S_{n}(f)} \,,
\ee
for stationary Gaussian noise with noise PSD $S_{n}$, the likelihood $\calL = p(d | \bm{\theta})$ for a given gravitational wave channel takes the form
\be\label{eq:deflnL}
	\ln \calL = -\frac{1}{2} \left( h(\bm{\theta}) - d | h(\bm{\theta}) - d \right) \,,
\ee
with the data stream being a superposition of the gravitational wave signal for the true parameters $\bm{\theta}_{0}$ and the noise realization in the experiment, $d = s(\bm{\theta}_{0}) + n$. The posterior distribution for the physical parameters $\bm{\theta}$ given the observed data $d$ is then
\be\label{eq:defBayes}
	p\left( \bm{\theta} | d \right) = \frac{p(d | \bm{\theta})p(\bm{\theta})}{p(d)}\,,
\ee
with $p(\bm{\theta})$ the priors on the parameters, and $p(d)$ the evidence that we will ignore here as it enters as a multiplicative constant in the posterior.

@@

\begin{figure*}
  \centering
  \includegraphics[width=.7\linewidth]{../plots/plot_tdia_hm.pdf}
  \caption{\SM{TODO: make figure smaller, single-column.}}
  \label{fig:FDHMSMBHCase9}
\end{figure*}

@@

\subsection{Time and frequency accumulation of SNR}
\label{sec:timefreqSNR}

-- SNR accumulation with time or frequency: inspiral/MRD balance across mass range

\subsection{A non-degenerate case}
\label{sec:SMBHPEnondeg}

@@

\subsection{Accumulating signal with time}
\label{sec:SMBHPEacctime}

@@

\subsection{SNR dependence of parameter estimation}
\label{sec:SMBHPEdistance}

-- Move to Appendix ?

@@

\begin{figure*}
  \centering
  \includegraphics[width=.8\linewidth]{../plots/corner_smbh_case9_simplelike_ptmcmc_bambi.png}
  \caption{.}
  \label{fig:PEsmbh22hmSimpleLikeCase9}
\end{figure*}

\begin{figure}
  \centering
  \begin{minipage}{.49\linewidth}
  \includegraphics[width=.99\linewidth]{../plots/corner_skydown_smbh_case9_22_ptmcmc_bambi_withdegen.png}
  \end{minipage}
  \begin{minipage}{.49\linewidth}
  \includegraphics[width=.99\linewidth]{../plots/corner_skydown_smbh_case9_simplelike_ptmcmc_bambi_withdegen.png}
  \end{minipage}
  \caption{.}
  \label{fig:Skydownsmbh22hmSimpleLikeCase9}
\end{figure}

@@

Under all these simplifying assumptions, the likelihood function becomes purely extrinsic, and takes the form of trivial geometric factors multiplying a constant overlap that can be precomputed. Using the subscript $\mathrm{inj}$ for the injected signal, the likelihood~\eqref{eq:defliklihood} is
\be
	\ln \calL = -\frac{1}{2} \left( \tilde{a} - \tilde{a}_{\rm inj} | \tilde{a} - \tilde{a}_{\rm inj} \right) -\frac{1}{2} \left( \tilde{e} - \tilde{e}_{\rm inj} | \tilde{e} - \tilde{e}_{\rm inj} \right) \,.
\ee
From the low-frequency response \eqref{eq:aelowfmodes}, pinning the time-at-LISA so that we ca ignore the factor $\exp [2 i \pi f k\cdot p_{0}]$ as a pure constant corresponding to a redefinition of time, we obtain 
\begin{align}
	\ln \calL &= -\frac{1}{2} \sum_{\ell m} \sum_{\ell' m'}   \left[ \left( s_{a}^{\ell m} - s_{a, \mathrm{inj}}^{\ell m}\right) \left( s_{a}^{\ell' m'} - s_{a, \mathrm{inj}}^{\ell' m'}\right)^{*} \right. \nn\\
	& \qquad \left. + \left( s_{e}^{\ell m} - s_{e, \mathrm{inj}}^{\ell m}\right) \left( s_{e}^{\ell' m'} - s_{e, \mathrm{inj}}^{\ell' m'}\right)^{*} \right] \left\langle \ell m | \ell' m' \right\rangle \,,
\end{align}
where we introduced the following extrinsic signal functions
\be\label{eq:defsaelm}
	s_{a,e}^{\ell m} = \frac{1}{d} \frac{2}{3L} D_{a,e} : P_{\ell m} \,,
\ee
with $d = D / D^{\rm inj}$ the dimensionless ratio of luminosity distances. The prefactors $\left\langle \ell m | \ell' m' \right\rangle$ are pure normalization constants, and represent mode overlaps as
\be
	\left\langle \ell m | \ell' m' \right\rangle = 4\mathrm{Re} \int \frac{df}{S_{n}^{a,e}} \left( 3 \pi f L\right)^{2} \tilde{h}_{\ell m} \tilde{h}_{\ell' m'}^{*} \,.
\ee
Here, the noise PSD $S_{n}^{a,e}$ is given by~\eqref{} and is identical between the channels $a$ and $e$. Having fixed the masses and the time, these overlaps are to be computed with the modes $\tilde{h}_{\ell m}^{\rm inj}$ of the injected signal.

In the case where we only include the dominant harmonic $h_{22}$, the functions~\eqref{eq:defsaelm} take the explicit form
\begin{align}
	s_{a,e}^{22} &= \frac{1}{4d} \sqrt{\frac{5}{\pi}} \cos^{4}\frac{\iota}{2} e^{2i(\varphi-\psiL)} \left( F_{a,e}^{+} + i F_{a,e}^{\times} \right) \\
	&+ \frac{1}{4d} \sqrt{\frac{5}{\pi}} \sin^{4}\frac{\iota}{2} e^{2i(\varphi+\psiL)} \left( F_{a,e}^{+} - i F_{a,e}^{\times} \right) \,,
\end{align}
with the pattern functions for the channels $a$ and $e$ defined in \eqref{eq:defFapcFepc}-\eqref{eq:FapcFepc}.

@@

We can now explain the degenerate features we found as certain limits in this simplified likelihood. In the face-on or face-off limit $\iota \rightarrow 0$, $|t_{\iota}| \ll 1$. This amounts to the $(\ell,m)=(2,2)$ contribution dominating the $(\ell, m)=(2,-2)$ contribution. Neglecting the terms in~\eqref{eq:sigmapm} that are quartic in $t_{\iota}$, we have
\bsub\label{eq:sigmapmapprox}
\begin{align}
	\sigma_{+} &\simeq \rho e^{2i(\varphi - \psiL)} t_{\theta}^{4} e^{-2i \lambda_{a}} \,, \\
	\sigma_{-} &\simeq \rho e^{2i(\varphi - \psiL)} e^{2i \lambda_{a}} \,,
\end{align}
\esub
Defining the ratio
\be
	r = \frac{\sigma_{+}}{\sigma_{-}} \,,
\ee
we see that finding $(\lambda_{a}^{*}, \theta_{L}^{*})$ verifying
\be\label{eq:conditionsky}
	\frac{\sigma_{+}}{\sigma_{-}} (\lambda_{a}^{*}, \theta_{L}^{*}) = r_{\rm inj} \,,
\ee
with the approximation~\eqref{eq:sigmapmapprox} for the left-hand side, guarantees that we can then match both $\sigma_{+}$ and $\sigma_{-}$ of the injection with the condition
\be\label{eq:conditionprefactor}
	\rho e^{2i(\varphi - \psiL)} e^{2i \lambda_{a}^{*}} = \sigma_{-}^{\rm inj}\,.
\ee
In the limit $|t_{\iota}| \ll 1$,
\be
	\frac{\sigma_{+}}{\sigma_{-}} \simeq t_{\theta}^{4} e^{-4i\lambda_{a}} \,,
\ee
so that~\eqref{eq:conditionsky} is verified for special positions in the sky at (coming back to the original variables $(\lambdaL, \betaL)$)
\bsub\label{eq:soldegensky}
\begin{align}
	\beta_{L}^{*} &= \frac{\pi}{2} - 2 \mathrm{Arctan} \; \left| r_{\rm inj}\right|^{1/4} \,, \\
	\lambda_{L}^{*} &= \frac{\pi}{6} -\frac{1}{4} \mathrm{Arg} \; r_{\rm inj} \;\mathrm{mod}\; \frac{\pi}{2} \,.
\end{align}
\esub
The other condition gives then a single phase constraint for $\varphi - \psiL$, and a single amplitude constraint for $\rho(d, \iota, \theta_{L}^{*})$. The limit $\iota \rightarrow \pi$ (where the $(2,-2)$ contribution dominates the $(2,2)$) is very similar, as can be seen most easily by the change of variable $\iota \rightarrow \pi - \iota$, with $-\psiL$ and $\pi - \theta_{L}$ playing the roles of $\psiL$ and $\theta_{L}$.

@@

We can further find the same mechanism, with a few differences, by exchanging the roles of $(\iota, \varphi)$ and $(\theta_{L}, \lambda_{a})$, as follows. In the limit $\theta_{L} \rightarrow 0$ with $|t_{\theta}| \ll 1$, we have this time
\be
	\frac{\sigma_{+}}{\sigma_{-}} \simeq t_{\iota}^{4} e^{-4i(\lambda_{a} - \psi_{L})} \,,
\ee
which gives in the same way a degeneracy if
\bsub\label{eq:soldegeniotaphi}
\begin{align}
	\iota^{*} &= 2 \mathrm{Arctan} \; \left| r_{\rm inj}\right|^{1/4} \,, \\
	(\lambda_{a} - \psiL)^{*} &= -\frac{1}{4} \mathrm{Arg} \; r_{\rm inj} \;\mathrm{mod}\; \frac{\pi}{2} \,,
\end{align}
\esub
with a condition on prefactors that is now concerning $\rho e^{2 i \varphi}$. Thus, we find again a degenerate pattern, with the difference that \eqref{eq:soldegeniotaphi} imposes a condition on the pair $\lambda_{a} - \psiL$ instead of $\lambda_{a}$, and the extra condition imposes a condition on $\varphi$ instead of the pair $\varphi - \psiL$.

We illustrate our findings by running a Bayesian parameter estimation for the simplified likelihood. The results are shown in Fig.~\ref{fig:PEsmbh22hmSimpleLikeCase9}, with the two samplers.

>>>>

@@

Introducing the standard matched-filter inner product as
\be\label{eq:definnerproduct}
	( \tilde{a} | \tilde{b} ) = 4 \mathrm{Re} \int_{0}^{+\infty} df \; \frac{\tilde{a}(f) \tilde{b}^{*}(f)}{S_{n}(f)} \,,
\ee
for stationary Gaussian noise with noise PSD $S_{n}$, the likelihood $\calL = p(d | \bm{\theta})$ for a given gravitational wave channel takes the form
\be\label{eq:deflnL}
	\ln \calL = -\frac{1}{2} \left( h(\bm{\theta}) - d | h(\bm{\theta}) - d \right) \,,
\ee
with the data stream being a superposition of the gravitational wave signal for the true parameters $\bm{\theta}_{0}$ and the noise realization in the experiment, $d = s(\bm{\theta}_{0}) + n$. The posterior distribution for the physical parameters $\bm{\theta}$ given the observed data $d$ is then
\be\label{eq:defBayes}
	p\left( \bm{\theta} | d \right) = \frac{p(d | \bm{\theta})p(\bm{\theta})}{p(d)}\,,
\ee
with $p(\bm{\theta})$ the priors on the parameters, and $p(d)$ the evidence that we will ignore here as it enters as a multiplicative constant in the posterior.

@@

\begin{figure}
  \centering
  \includegraphics[width=.99\linewidth]{../plots/plot_hc_tdia_hm.pdf}
  \caption{.}
  \label{fig:FDHMSMBHCase9}
\end{figure}

-- add plot for SNR accumulation with time or frequency: inspiral/MRD balance across mass range

@@

\subsection{Parameter estimation}
\label{sec:SMBHPE}

@@

\subsection{Accumulation of signal with time}
\label{sec:SMBHPEacctime}

@@

@@

\begin{figure*}
  \centering
  \includegraphics[width=.8\linewidth]{../plots/corner_smbh_case9_simplelike_ptmcmc_bambi.png}
  \caption{Posterior distributions obtained with the two samplers \texttt{PTMCMC} (blue) and $\texttt{MultiNest}$ (green) for the simplified likelihood~\eqref{eq:simplelikelihoodhm}, using only the dominant mode $(\ell, m) = (2,2)$. Masses and time are pinned to the injected value, and the motion of the detector is neglected. We also superimpose (red) the contours obtained with \texttt{MultiNest} by including all harmonics $(2,2)$, $(2,1)$, $(3,3)$, $(4,4)$, $(5,5)$.}
  \label{fig:PEsmbh22hmSimpleLikeCase9}
\end{figure*}

\begin{figure}
  \centering
  \begin{minipage}{.49\linewidth}
  \includegraphics[width=.99\linewidth]{../plots/corner_skydown_smbh_case9_22_ptmcmc_bambi_withdegen.png}
  \end{minipage}
  \begin{minipage}{.49\linewidth}
  \includegraphics[width=.99\linewidth]{../plots/corner_skydown_smbh_case9_simplelike_ptmcmc_bambi_withdegen.png}
  \end{minipage}
  \caption{Sky position posterior in presence of degeneracies, zooming close to the injected value. The left panel shows the complete likelihood, while the right panel shows the simplified likelihood~\eqref{eq:simplelikelihoodhm} with a motionless LISA and fixed masses and time. The results of \texttt{PTMCMC} (blue) and \texttt{MultiNest} are superimposed. The black cross is the injected value, and the grey cross is the analytic prediction~\eqref{eq:soldegensky} for the offset peak created by degeneracies.}
  \label{fig:Skydownsmbh22hmSimpleLikeCase9}
\end{figure}

@@

Under all these simplifying assumptions, the likelihood function becomes purely extrinsic, and takes the form of trivial geometric factors multiplying a constant overlap that can be precomputed. Using the subscript $\mathrm{inj}$ for the injected signal, the likelihood~\eqref{eq:defliklihood} is
\be
	\ln \calL = -\frac{1}{2} \left( \tilde{a} - \tilde{a}_{\rm inj} | \tilde{a} - \tilde{a}_{\rm inj} \right) -\frac{1}{2} \left( \tilde{e} - \tilde{e}_{\rm inj} | \tilde{e} - \tilde{e}_{\rm inj} \right) \,.
\ee
From the low-frequency response \eqref{eq:aelowfmodes}, pinning the time-at-LISA so that we ca ignore the factor $\exp [2 i \pi f k\cdot p_{0}]$ as a pure constant corresponding to a redefinition of time, we obtain 
\begin{align}\label{eq:simplelikelihoodhm}
	\ln \calL &= -\frac{1}{2} \sum_{\ell m} \sum_{\ell' m'}   \left[ \left( s_{a}^{\ell m} - s_{a, \mathrm{inj}}^{\ell m}\right) \left( s_{a}^{\ell' m'} - s_{a, \mathrm{inj}}^{\ell' m'}\right)^{*} \right. \nn\\
	& \qquad \left. + \left( s_{e}^{\ell m} - s_{e, \mathrm{inj}}^{\ell m}\right) \left( s_{e}^{\ell' m'} - s_{e, \mathrm{inj}}^{\ell' m'}\right)^{*} \right] \left\langle \ell m | \ell' m' \right\rangle \,,
\end{align}
where we introduced the following extrinsic signal functions
\be\label{eq:defsaelm}
	s_{a,e}^{\ell m} = \frac{1}{d} \frac{2}{3L} D_{a,e} : P_{\ell m} \,,
\ee
with $d = D / D^{\rm inj}$ the dimensionless ratio of luminosity distances. The prefactors $\left\langle \ell m | \ell' m' \right\rangle$ are pure normalization constants, and represent mode overlaps as
\be\label{eq:innerproductlmlpmp}
	\left\langle \ell m | \ell' m' \right\rangle = 4\mathrm{Re} \int \frac{df}{S_{n}^{a,e}} \left( 3 \pi f L\right)^{2} \tilde{h}_{\ell m} \tilde{h}_{\ell' m'}^{*} \,.
\ee
Here, the noise PSD $S_{n}^{a,e}$ is given by~\eqref{} and is identical between the channels $a$ and $e$. Having fixed the masses and the time, these overlaps are to be computed with the modes $\tilde{h}_{\ell m}^{\rm inj}$ of the injected signal.

In the case where we only include the dominant harmonic $h_{22}$, the likelihood~\eqref{eq:simplelikelihoodhm} simplifies to
\be\label{eq:simplelikelihood22}
	\ln \calL = -\frac{1}{2} \left\langle 22 | 22 \right\rangle  \left[ \left| s_{a}^{22} - s_{a, \mathrm{inj}}^{22} \right|^{2} + \left| s_{e}^{22} - s_{e, \mathrm{inj}}^{22}\right|^{2} \right] \,.
\ee
The functions~\eqref{eq:defsaelm} take the explicit form
\begin{align}
	s_{a,e}^{22} &= \frac{1}{4d} \sqrt{\frac{5}{\pi}} \cos^{4}\frac{\iota}{2} e^{2i(\varphi-\psiL)} \left( F_{a,e}^{+} + i F_{a,e}^{\times} \right) \\
	&+ \frac{1}{4d} \sqrt{\frac{5}{\pi}} \sin^{4}\frac{\iota}{2} e^{2i(\varphi+\psiL)} \left( F_{a,e}^{+} - i F_{a,e}^{\times} \right) \,,
\end{align}
with the pattern functions for the channels $a$ and $e$ defined in \eqref{eq:defFapcFepc}-\eqref{eq:FapcFepc}.

@@

We can now explain the degenerate features we found as certain limits in this simplified likelihood. In the face-on or face-off limit $\iota \rightarrow 0$, $|t_{\iota}| \ll 1$. This amounts to the $(\ell,m)=(2,2)$ contribution dominating the $(\ell, m)=(2,-2)$ contribution. Neglecting the terms in~\eqref{eq:sigmapm} that are quartic in $t_{\iota}$, we have
\bsub\label{eq:sigmapmapprox}
\begin{align}
	\sigma_{+} &\simeq \rho e^{2i(\varphi - \psiL)} t_{\theta}^{4} e^{-2i \lambda_{a}} \,, \\
	\sigma_{-} &\simeq \rho e^{2i(\varphi - \psiL)} e^{2i \lambda_{a}} \,,
\end{align}
\esub
Defining the ratio
\be
	r = \frac{\sigma_{+}}{\sigma_{-}} \,,
\ee
we see that finding $(\lambda_{a}^{*}, \theta_{L}^{*})$ verifying
\be\label{eq:degenconditionsky}
	\frac{\sigma_{+}}{\sigma_{-}} (\lambda_{a}^{*}, \theta_{L}^{*}) = r_{\rm inj} \,,
\ee
with the approximation~\eqref{eq:sigmapmapprox} for the left-hand side, guarantees that we can then match both $\sigma_{+}$ and $\sigma_{-}$ of the injection with the condition
\be\label{eq:degenconditionprefactor}
	\rho e^{2i(\varphi - \psiL)} e^{2i \lambda_{a}^{*}} = \sigma_{-}^{\rm inj}\,.
\ee
In the limit $|t_{\iota}| \ll 1$,
\be
	\frac{\sigma_{+}}{\sigma_{-}} \simeq t_{\theta}^{4} e^{-4i\lambda_{a}} \,,
\ee
so that~\eqref{eq:degenconditionsky} is verified for special positions in the sky at (coming back to the original variables $(\lambdaL, \betaL)$)
\bsub\label{eq:soldegensky}
\begin{align}
	\beta_{L}^{*} &= \frac{\pi}{2} - 2 \mathrm{Arctan} \; \left| r_{\rm inj}\right|^{1/4} \,, \\
	\lambda_{L}^{*} &= \frac{\pi}{6} -\frac{1}{4} \mathrm{Arg} \; r_{\rm inj} \;\mathrm{mod}\; \frac{\pi}{2} \,.
\end{align}
\esub
The other condition~\eqref{eq:conditionprefactor} gives then a single phase constraint for $\varphi - \psiL$, and a single amplitude constraint for $\rho(d, \iota, \theta_{L}^{*})$. The limit $\iota \rightarrow \pi$ (where the $(2,-2)$ contribution dominates the $(2,2)$) is very similar, as can be seen most easily by the change of variable $\iota \rightarrow \pi - \iota$, with $-\psiL$ and $\pi - \theta_{L}$ playing the roles of $\psiL$ and $\theta_{L}$.

@@

We can further find the same mechanism, with a few differences, by exchanging the roles of $(\iota, \varphi)$ and $(\theta_{L}, \lambda_{a})$, as follows. In the limit $\theta_{L} \rightarrow 0$ with $|t_{\theta}| \ll 1$, we have this time
\be
	\frac{\sigma_{+}}{\sigma_{-}} \simeq t_{\iota}^{4} e^{-4i(\lambda_{a} - \psi_{L})} \,,
\ee
which gives in the same way a degeneracy if
\bsub\label{eq:soldegeniotaphi}
\begin{align}
	\iota^{*} &= 2 \mathrm{Arctan} \; \left| r_{\rm inj}\right|^{1/4} \,, \\
	(\lambda_{a} - \psiL)^{*} &= -\frac{1}{4} \mathrm{Arg} \; r_{\rm inj} \;\mathrm{mod}\; \frac{\pi}{2} \,,
\end{align}
\esub
with a condition on $\rho$ and $\varphi$ that is now
\be
	\rho e^{2i\varphi} e^{2i (\lambda_{a}^{*} - \psi_{L}^{*})} = \sigma_{-}^{\rm inj} \,.
\ee
Thus, we find again a degenerate pattern, with the difference that \eqref{eq:soldegeniotaphi} imposes a condition on the pair $\lambda_{a} - \psiL$ instead of $\lambda_{a}$, and the extra condition imposes a condition on $\varphi$ instead of the pair $\varphi - \psiL$.
